Unsupervised

### 2023-12-14 EDA.Rmd
hi

---
title: "EDA"
author: "Mihir Kulkarni, Nithika Menon"
date: "2023-12-12"
output: html_document
---

@@

@@

This question investigates the effectiveness of the drug D-penicillamine compared to a placebo, considering the stage of PBC. By analyzing survival rates across different disease stages and treatment types, we can assess the drug's effectiveness at various disease stages. This information is vital for clinicians to make informed decisions about treatment plans and for researchers to understand the drug's impact.

@@

### Sex Distribution of Patients
```{r}
library(ggplot2)
library(dplyr)

cirrhosis$Sex <- factor(ifelse(cirrhosis$Sex == 0, "Female", "Male"))

# Pie Chart
ggplot(sex_data, aes(x="", y=Percentage, fill=Sex)) +
    geom_bar(stat="identity", width=1) +
    coord_polar("y", start=0) +
    scale_fill_brewer(palette="Pastel1") +
    theme_minimal() +
    theme(axis.line = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid = element_blank(),
          plot.title = element_text(size=14, face="bold")) +
    labs(fill="Sex", title="Distribution of Sex in PBC Patients")

### Age Distribution of Patients

```{r}
# Converting age from days to years for better readability
cirrhosis$Age_Years <- cirrhosis$Age / 365.25

### might have to mutate data below to filter out alive ppl but get the gist
```{r}

ggplot(data = cirrhosis, aes (Drug, Stage, fill = N_Days))  + geom_tile() + scale_fill_distiller(palette = "RdPu")

### Individuals with Edema and if they were on the drug
```{r}

library(ggmosaic)
cirrhosis %>%
  ggplot() +
  geom_mosaic(aes( x = product(Edema), fill = Drug))
#ggplot(data = cirrhosis) + geom_mosaic(aes(x = Drug , fill = Status))

```
Looking at the dist. of stages
```{r}

cirrhosis %>%
      ggplot(aes(x = Stage)) + geom_bar() # need color

```{r}
cirrhosis %>%
      ggplot(aes(x = Cholesterol, y = Bilirubin)) + geom_line()

### Comparison of Edema Status on Patient Outcomes

Edema is a common complication in liver diseases and can significantly impact patient quality of life and survival. Understanding how different levels of edema (none, controlled by diuretics, or persistent despite treatment) affect survival can inform patient management strategies and highlight the need for aggressive interventions in certain cases.

>>>>

---
title: "EDA"
author: "Mihir Kulkarni, Nithika Menon"
date: "2023-12-12"
output: html_notebook
---

@@

# EDA

## Exploring the demographics of the patients

### Sex Ratio of Patients
```{r}

#cirrhosis$Sex <- factor(ifelse(cirrhosis$Sex == 0, "Female", "Male"))

# Pie Chart
ggplot(sex_data, aes(x="", y=Percentage, fill=Sex)) +
    geom_bar(stat="identity", width=1) +
    coord_polar("y", start=0) +
    scale_fill_brewer(palette="Pastel1") +
    theme_minimal() +
    theme(axis.line = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid = element_blank(),
          plot.title = element_text(size=14, face="bold")) +
    labs(fill="Sex", title=" Sex Amongst Patients")

### Age Distribution of Patients
```{r}
# Converting age from days to years for better readability
cirrhosis$Age_Years <- cirrhosis$Age / 365.25

## Looking at the data itself

### NA counts amongst our variables
```{r}
count <-  data.frame(NA_CountsPerVar = colSums(is.na(cirrhosis)))
kable(count)
```
An observation is that many of the variables have 106 NAs. This indicates that a good fraction [106 patients] were equally tracked and measured in a less extensive way.

## Distribution of Stages of Biliary Cholangitis
```{r}

cirrhosis %>%
      ggplot(aes(x = Stage)) + geom_bar(fill = "#097969") # need color

```
This distribution is left skewed and not symmetric. Most patients have stage 3 and then 4 of Biliary Cholangitis.

## Exploring some interactions of variables

### How does prescense of a D-penicillamine with stage of Biliary Cholangitis impact number of days till death?
```{r}

ggplot(data = cirrhosis, aes (Drug, Stage, fill = N_Days))+ geom_tile() + scale_fill_distiller(palette = "RdPu", trans = "reverse")

```
This heat map shows that being a placebo in stage 1 gives you a greater amount of days till death in this sample of patients.

### Individuals with Edema, and if they were on the drug
```{r}

hi = c("#40B5AD", "#009E60", "#9FE2BF")
library(ggmosaic)
cirrhosis %>%
  ggplot() +
  geom_mosaic(aes( x = product(Edema), fill = Drug)) + scale_fill_manual(values = hi)
```

Edema is swelling due to too much liquid trapped in the body's tissues. It's common complication in liver diseases and can significantly impact patient quality of life and survival. Understanding how different levels of edema (none, controlled by diuretics, or persistent despite treatment) affect survival can inform patient management strategies and highlight the need for aggressive interventions in certain cases. From the graph we can see a even split between the placebo and drug for Edema status. There is no NA values for Edema persistent despite diuretics. We cannot  directly compare it the splits for other statuses given their NA values if known could change the look of the graph.

### mihir choose your preffered graph and explain

@@

This question investigates the effectiveness of the drug D-penicillamine compared to a placebo, considering the stage of PBC. By analyzing survival rates across different disease stages and treatment types, we can assess the drug's effectiveness at various disease stages. This information is vital for clinicians to make informed decisions about treatment plans and for researchers to understand the drug's impact. (LOLL this is what the study acc sought to discover)

@@

### 2023-12-14 Final Project.Rmd
hi

@@

### Q2: Are there factors that can help distinguish different stages of Biliary Cholangitis not currently included in medical definition?

@@

- Data Preprocessing

```{r}
#load in the data
library(tidyverse)
#load data
cirrhosis <- read_csv("cirrhosis.csv")
head(cirrhosis)
```

# Pre-process the data
```{r}
#encode the sex, ascites, hepatomegaly, spiders, edema into binary variables
cirrhosis$Sex <- ifelse(cirrhosis$Sex == "F", 0, 1)
cirrhosis$Ascites <- ifelse(cirrhosis$Ascites == "Y", 1, 0)
cirrhosis$Hepatomegaly <- ifelse(cirrhosis$Hepatomegaly == "Y", 1, 0)
cirrhosis$Spiders <- ifelse(cirrhosis$Spiders == "Y", 1, 0)
#factorize the stage
cirrhosis$Stage <- factor(cirrhosis$Stage)
cirrhosis$Status <- factor(cirrhosis$Status)
cirrhosis$Drug <- factor(cirrhosis$Drug)
cirrhosis$Sex <- factor(cirrhosis$Sex)
cirrhosis$Ascites <- factor(cirrhosis$Ascites)
cirrhosis$Hepatomegaly <- factor(cirrhosis$Hepatomegaly)
cirrhosis$Spiders <- factor(cirrhosis$Spiders)
cirrhosis$Edema <- factor(cirrhosis$Edema)
#need to do more
#Drop the first column
cirrhosis <- cirrhosis[,-1]
```

```{r}
count <- colSums(is.na(cirrhosis))

kable(count)
```
- Engage in Exploratory Data Analysis

>>>>

# ASK MIHIR TO PUT IN IMAGES HERE !!!

@@

### Q2: Are there factors not currently included in medical definition that can help distinguish different stages of Biliary Cholangitis?

@@

### Load in Packages
```{r}

library(survival)
library(ggplot2)
library(survival)
library(survminer)
library(kableExtra)
library(factoextra)

### Data Processing
```{r}
#load in the data
library(tidyverse)
#load data
cirrhosis <- read_csv("cirrhosis.csv")
head(cirrhosis)
```

### Pre-process the data
```{r}
#encode the sex, ascites, hepatomegaly, spiders, edema into binary variables
cirrhosis$Sex <- ifelse(cirrhosis$Sex == "F", 0, 1)
cirrhosis$Ascites <- ifelse(cirrhosis$Ascites == "Y", 1, 0)
cirrhosis$Hepatomegaly <- ifelse(cirrhosis$Hepatomegaly == "Y", 1, 0)
cirrhosis$Spiders <- ifelse(cirrhosis$Spiders == "Y", 1, 0)
#factorize the stage
cirrhosis$Stage <- factor(cirrhosis$Stage)
cirrhosis$Status <- factor(cirrhosis$Status)
cirrhosis$Drug <- factor(cirrhosis$Drug)
cirrhosis$Sex <- factor(cirrhosis$Sex)
#cirrhosis$Ascites <- factor(cirrhosis$Ascites)
#cirrhosis$Hepatomegaly <- factor(cirrhosis$Hepatomegaly)
#cirrhosis$Spiders <- factor(cirrhosis$Spiders)
cirrhosis$Edema <- factor(cirrhosis$Edema)
cirrhosis$Sex <- factor(ifelse(cirrhosis$Sex == 0, "Female", "Male"))

#Drop the first column
cirrhosis <- cirrhosis[,-1]
```

- Engage in Exploratory Data Analysis

### 2023-12-14 Unsupervised.Rmd
Added basic code - EDA, importing data, tree and logistic regression
hi

---
title: "Regression-based"
author: "Mihir Kulkarni"
date: "2023-12-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

>>>>

---
title: "Unsupervised"
author: "Mihir Kulkarni"
date: "2023-12-12"
output: html_notebook
---

# Unsupevised Learning

 In this section, we sought to if there are sub-collections of patients. We expect sub-collections to mimic medical definitions of the stages of Biliary Cirrhosis but are curious to see if clustering can reveal some newer observations.

### Stages of Biliary Cirrhosis

Stage 1: There's inflammation and damange to the walls of medium-sized bile ducts.

Stage 2: There's blockage of the small bile ducts.

Stage 3: This stage marks the beginning of scarring.

Stage 4: Cirrhosis has developed. This permanent, severe, scarring and damange to the liver.

Source of information: https://www.healthline.com/health/primary-biliary-cirrhosis#stages

#do the ml, take about observational stages vs now quantify, look into the different variables and what they mean.

### Data Wrangling
```{r}

cirrhosisCluster <- cirrhosis

cirrhosisCluster <- cirrhosisCluster %>%                 
                                  select(-Stage)

cirrhosisCluster <- na.omit(cirrhosisCluster) #cannot have NA values in clustering

cirrhosisCluster$Age_Years <- round(cirrhosisCluster$Age_Years) #round age to whole numbers

#reode sex... recode others later if need be #female is 0

#cirrhosisCluster  <- cirrhosisCluster %>%
#  select( -c(Status, Drug, Edema)) %>%

# mutate(Sex = ifelse(Sex == 'Female', 0, 1))

  cirrhosisCluster  <- cirrhosisCluster %>%
    mutate(Sex = ifelse(Sex == 'Female', 0, 1)) %>%
          mutate(Transplant = ifelse(Status == "CL", 1, 0)) %>%
              mutate(Status = ifelse(Status %in% c('C', 'CL'), 0 , 1)) %>%
                    mutate(Drug = ifelse(Drug == "D-penicillamine", 0, 1)) %>%
                        mutate(EdemaDiurectics = ifelse(Edema %in% c('S', 'Y'), 1, 0)) %>%
                              mutate(NoEdemaORD = ifelse(Edema == 'N' , 1, 0)) %>%
                                        mutate(EdemaANDD = ifelse(Edema == "Y", 1, 0)) %>%
                                                mutate(EdemaORD = ifelse(Edema == "S", 1, 0)) %>%
                                                                              select(-Edema)

#Data is already factored

#mutate(Sex = ifelse(Sex == 'Female', 0, 1)) %>%

   #mutate(Status = ifelse(Status %in% c('C', 'CL'), 0 , 1)) %>%
          #    mutate(Drug = ifelse(Drug == "D-penicillamine", 0, 1))
```

#make column yes no edema
#then yes no under treatment

#same for tranplant stuff

```{r}
distCirrhosis <- dist(
  x = cirrhosisCluster,
  method = "euclidean"
)

```

### A mixed hierarchial and non-hierarchial

```{r}

hybridCirrhosis <- hkmeans(
  x = cirrhosisCluster,
  k = 4,
  hc.metric = "euclidean",
  hc.method = "ward.D",
  iter.max = 10
)

```

### Visualize Results

### Making a color palette
```{r}

StagesPalette <- c("#AA336A", "#770737", "#40B5AD", "#009E60")

```

```{r}

## MAKE A NEW PALLETE TO VISUALIZE

# Plot the initial dendrogram for hybrid approach ----
set.seed(380)
hkmeans_tree(
  hkmeans = hybridCirrhosis,
  rect.col = StagesPalette,
  cex = 0.4,
  main = "Initial Hierarchical Clusters"
)

```

```{r}
# Plot the final dendrogram for hybrid approach ----
# library(factoextra)
fviz_dend(
  x = hybridCirrhosis,
  cex = 0.4,
  palette = StagesPalette,
  rect = FALSE,
  horiz = TRUE,
  repel = TRUE,
  main = "Final Dendrogram"
)

```

```{r}

# Plot the final clustering for hybrid approach ----
# library(factoextra)
fviz_cluster(
  object = hybridCirrhosis,
  stand = FALSE,
  geom = "point",
  main = "Hybrid Cluster Plot"
) +
  scale_color_manual(values = StagesPalette) +
  scale_fill_manual(values = StagesPalette) +
  theme_bw()

```
